High-Order Accurate Finite Difference Schemes \\ for Systems of PDEs via Optimal Time Step Selection

### 2008-08-10 papers/OTSPDE-Systems/OTSPDE-Systems.tex
OTSPDE: * Checked in place holders for upcoming papers. * Updated information on potential journals to submit to.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Manuscript started: 2007/09/01
% First completed draft: 2008/05/12, Vitamin D, Inc.
% Submission draft: 2008/06/08, Vitamin D, Inc.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\title{High-Order Accurate Finite Difference Schemes \\
       Via Optimal Time Step Selection}

@@

\begin{abstract}
In this article, we present a novel technique for computing high-order
accurate numerical solutions to time dependent partial differential 
equations using only formally low-order finite difference stencils and
time integration schemes.  We achieve this high-order accuracy in a simple 
manner by optimally choosing the time step for the low-order finite 
difference scheme (and possibly adding a few correction terms).  Through 
straightforward numerical analysis arguments, we explain the origin of the 
boost in accuracy and estimate the computational cost of the high-order 
methods obtained via optimal time step (OTS) selection.  We demonstrate the 
utility of OTS selection on a wide array of classical PDEs in one and more 
space dimensions and find that it is applicable to linear and semilinear PDEs 
on both regular and irregular domains. 
\end{abstract}

\begin{keywords}
optimal time step, finite difference schemes, high-order accurate numerical 
methods, time dependent PDEs
\end{keywords}

\begin{AMS}
65-02, 65M06, 65M12, 65M20
\end{AMS}

@@

There are many high-order accurate methods that are simple combinations or 
modifications of low-order schemes.  A few well-known examples include 
Richardson extrapolation~\cite{richardson_1927, heath_book}, the modified 
nine-point formula for the 2D Poisson equation~\cite{iserles_book}, and the 
Lax-Wendroff method for the advection equation~\cite{leveque_book_1992, 
leveque_book_2002, gko_book}.  Cancellation of higher-order errors is a key 
principle underlying the superior accuracy of these methods.

More commonly, high-order accurate numerical methods are constructed by 
requiring that the discretization error possesses the desired formal order of 
accuracy for \emph{all} of the independent variables in the PDE.  While this 
constraint is certainly sufficient for high-order accuracy, it is by no means 
necessary.  This conventional approach for constructing numerical methods 
has its roots in the classical numerical analysis of PDEs, which tends to 
analyze the discretization of the terms in the PDE individually.  Within this 
framework, it is natural to seek high-order accuracy by ensuring that the 
numerical scheme is simultaneously high-order accurate for all independent 
variables.  Unfortunately, this common approach for developing and analyzing 
numerical methods is often suboptimal because it neglects the potential for 
achieving high-order accuracy using simple low-order schemes.

Perhaps the greatest weakness of the standard method for constructing 
high-order numerical methods is the use of ``big-O" notation, which hides 
the details of the discretization error.  The information lost in the 
``big-O" notation eliminates the possibility of reducing numerical error by 
taking advantage of relationships between the terms in the error.  The 
modified nine-point formula for the Poisson equation~\cite{iserles_book} 
and the Lax-Wendroff scheme for the advection equation~\cite{leveque_book_1992,
leveque_book_2002, gko_book} demonstrate the power of explicitly retaining 
higher-order terms in the discretization error when analyzing numerical 
schemes.  For both of these methods, examination of higher-order terms allows 
low-order methods to be transformed into high-order methods with only minor 
modifications to the numerical scheme.

The usual approach for developing numerical methods has another significant
shortcoming -- it places no value on the optimization of numerical parameters 
(\eg $\dx$ and $\dt$) for accuracy.  Again, ``big-O" notation is 
somewhat to blame for this situation because it gives the impression 
that the error depends on each numerical parameter independently.  Aside
from constraints needed for numerical stability, ``big-O" notation makes 
it appear as though any combination of numerical parameters is equally good 
as long as they are all ``small enough.''  For some problems, however, 
optimizing the numerical parameters can significantly increase the accuracy
of the numerical solution.  For example, in~\cite{zhao_2006}, Zhao showed 
that choosing an appropriate value for $\theta$ as a function of the time 
step in the theta-method\footnote{In~\cite{zhao_2006} refers to $\theta$ as 
the weight factor and the theta-method as the single-step trapezoidal 
method.}~\cite{iserles_book, gko_book} can decrease the error of finite 
element solutions to the diffusion equation.

In this paper, we explore a novel technique for constructing high-order 
finite difference methods for time dependent PDEs using only formally 
low-order stencils and time integration schemes.  The basic idea is that a 
carefully chosen time step (and possibly the addition of a few correction 
terms) can eliminate low-order terms in the discretization error, allowing a 
formally low-order method to deliver high-order accuracy.  The optimal time 
step is calculated in a straightforward manner by using knowledge of the 
PDE to eliminate the leading-order terms in the discretization error.  
Optimal time step selection combines the well-known approach of using 
the PDE to reduce numerical error with the less-common notion that optimal 
choice of numerical parameters can boost the effectiveness of algorithmically 
simple and computationally low-cost methods.

Using the PDE to reduce discretization error has a long 
tradition in numerical analysis.  Indeed, this idea underlies both the 
modified nine-point formula for the Poisson equation~\cite{iserles_book} and 
the Lax-Wendroff method for the linear advection 
equation~\cite{leveque_book_1992, leveque_book_2002, gko_book}.  
In the case of the modified nine-point formula for the Poisson equation, a 
simple analysis of the nine-point stencil for the Laplacian on a uniform grid 
shows that it is only second-order accurate, which might lead to the conclusion 
that the nine-point stencil has no advantage over the standard five-point 
stencil.  However, a more careful analysis of the discretization error in 
light of the PDE reveals that the leading-order error, which is proportional 
to the bilaplacian of the solution~\cite{iserles_book, patra_2005}, is directly 
related to the Laplacian of the source term in the Poisson equation.  Thus, 
the numerical scheme is easily boosted to fourth-order by simply adding a 
correction term that cancels out the leading-order error.   
Similarly, use of the PDE to eliminate higher-order discretization errors also 
plays a critical role in the Lax-Wendroff method because it allows the second 
derivative of the solution in time to be written in terms of the second 
derivative in space.  This observation transforms an unstable, first-order 
scheme based on a central difference approximation of 
$\partial/\partial x$ into a stable, second-order 
scheme~\cite{leveque_book_1992, leveque_book_2002}.

Optimal time step selection as a technique for improving the overall accuracy 
of numerical schemes for time dependent PDEs appears to be a novel idea.
Typically, the primary concern when choosing a time step is numerical 
stability with accuracy being a secondary consideration.  The idea that a 
clever choice of time step can be used to dramatically reduce the error is 
rarely considered.  In general, little research seems to have been done in
the area of optimizing the parameters of numerical methods for PDEs (\eg time 
step and grid spacing) to improve accuracy.  To the best of the author's 
knowledge, Zhao's method for choosing an optimal value of $\theta$ in the 
theta-method is the only work that attempts to improve accuracy by modifying 
numerical parameters.  While he focuses solely on the optimal choice of 
$\theta$ for a fixed time step, his approach can be viewed as a form of 
optimal time step selection by reversing the roles of $\theta$ and the time 
step.

It is worth spending a few moments to draw the distinction between optimal 
time step selection and traditional adaptive time stepping 
techniques~\cite{iserles_book, shampine_2005} that are used in the 
context of the method 
of lines~\cite{iserles_book, gko_book}.  On the surface, optimal time step 
selection may appear to be just a crude version of adaptive time step control.
However, the two methods are actually very different.  While both methods seek 
to reduce numerical error by controlling the time step, the numerical errors 
they control are fundamentally different.  Adaptive time stepping can only 
be used to reduce temporal discretization errors because it controls the 
errors that arise when solving the coupled system of ODEs for the 
semi-discrete approximation to the PDE; the spatial discretization errors 
are completely controlled by the finite difference scheme selected to 
approximate the spatial derivatives.  
In contrast, optimal time step selection simultaneously reduces 
\emph{both} the spatial and temporal discretization errors because it
uses information from the PDE to choose the time step.  Thus, optimal time 
step selection can potentially produce superior results compared with 
traditional adaptive time stepping approaches for integrating the 
semi-discrete equations associated with time dependent PDEs.  Another 
important distinction between the two techniques is that optimal time step 
selection uses a \emph{fixed} time step which makes it significantly less 
computationally complex than adaptive time stepping methods.

As a technique for constructing numerical methods, optimal time step selection 
has several desirable features.  Like any technique that leads to high-order 
methods, optimal time step selection produces schemes that greatly reduce the 
computational cost required to obtain a numerical solution of a desired 
accuracy.  However, its real power comes from the fact that it achieves
high-order accuracy while being based solely on very simple finite difference 
schemes.  For instance, we will see how optimal time step selection makes 
it possible to achieve 4th-order accuracy for the diffusion equation using only 
a second-order stencil for the Laplacian and forward Euler time integration.  
Moreover, high-order accuracy is not restricted to problems on simple, 
rectangular domains; irregular domains are straightforward to handle by
appropriately setting the values in ghost cells~\cite{gibou_2005, fedkiw_1999, 
osher_fedkiw_book}.  Optimal time step selection is even beneficial when 
using finite difference schemes to solve some nonlinear PDEs.  While there 
are certainly limitations to optimal time step selection, we will see that 
there are many problems where it is useful.

The remainder of this paper is organized as follows.  
In Section~\ref{sec:optimal_time_step_selection}, we begin by 
discussing the technique of optimal time step selection and 
presenting the analysis that explains how it boosts the order of accuracy 
for formally low-order numerical schemes.  In 
Section~\ref{sec:applications_1d}, we demonstrate the general utility of 
optimal time selection to problems in one space dimension by applying it to 
finite difference schemes for several classical linear and nonlinear PDEs. 
In Section~\ref{sec:applications_multidim}, we use optimal time step selection 
to improve the accuracy of simple finite difference schemes for the linear 
advection equation and diffusion equation in multiple space dimensions.  
To illustrate the ease with which high-order accuracy can be achieved for
problems on arbitrary domains, the 2D diffusion equation is solved on both 
the regular and irregular domains.  
For each PDE in Sections~\ref{sec:applications_1d} and 
\ref{sec:applications_multidim}, we provide the numerical analysis required 
to calculate the optimal time step size and order of accuracy.  
Finally, in Section~\ref{sec:summary}, we provide some concluding remarks.

\section{\label{sec:optimal_time_step_selection} 
         Optimal Time Step (OTS) Selection}
Optimal time step selection is not by itself a method for constructing 
finite difference schemes.  Rather it is a way to enhance to the performance
of existing finite difference schemes by carefully choosing the time step to 
eliminate low-order numerical errors.  The two basic ideas underlying 
optimal time step selection are (1) the PDE provides valuable insight into the 
discretization errors for finite difference schemes and (2) a judicious choice 
of time step can be used to eliminate the leading-order (and possibly 
higher-order) terms in the error.  In this section, we present and analyze 
the technique of optimal time step selection for finite difference schemes 
constructed to solve evolution equations of the form 
\beq
  \frac{\partial u}{\partial t} = 
    F\left(x, u(x), D u(x), D^2 u(x), \ldots \right), 
\eeq
where the right hand side is an arbitrary function of the solution $u$, its 
spatial derivatives, and the independent variable $x$.
Because this class of time dependent PDEs is so prevalent and important in 
applied mathematics, restricting our attention to problems of this form is 
not a serious limitation.

\subsection{\label{sec:ots_model_pde} 
            Optimal Time Step for a Model PDE}
Let us begin by considering a finite difference approximation for the
following time dependent PDE in one spatial dimension: 
\beq
  \frac{\partial u}{\partial t} = A \frac{\partial^n u}{\partial x^n},
  \label{eq:model_PDE}
\eeq
where $u$ is the solution to the PDE, $n$ is the spatial order of the 
PDE, and $A$ is a constant of the appropriate sign so that~(\ref{eq:model_PDE}) 
is well-posed.  Because we will be boosting the accuracy of the finite 
difference approximation by optimizing the time step, there is no reason to 
explicitly construct the scheme to be high-order.  In fact, using a high-order 
finite difference scheme only complicates the analysis required to calculate 
the optimal time step.

To compute the optimal time step, we first analyze the discretization errors 
(both spatial and temporal) for the finite difference approximation, 
explicitly keeping track of the leading-order terms.  In contrast to 
conventional error analysis, we do not sweep all of the errors under ``big-O'' 
notation.  Rather, we take advantage of the fact that derivation of the 
leading-order terms in the error is straightforward for finite difference
schemes through the use of Taylor series expansions.  Because each term in 
the error is proportional to a partial derivative of the solution, we can 
attempt to use the PDE to relate the leading-order terms to each other.  If we 
are fortunate (which we are for many finite difference schemes), this careful 
analysis yields a direct relationship between the leading-order spatial and 
temporal errors that allows them to be combined as a single term of the form:
\beq
  (L u) P(\dx, \dt) ,
  \label{eq:leading_order_error_model_PDE_general}
\eeq
where $L$ is a differential operator and $P$ is a polynomial in the numerical 
parameters $\dx$ and $\dt$.  By selecting the numerical parameters so that 
$P = 0$, we can completely eliminate the leading-order term in the 
discretization error.  Because time dependent PDEs are often solved by 
stepping in time, it is natural to choose the time step to be a function of 
the grid spacing.  The time step selected in this way is the
\emph{optimal time step} for the finite difference scheme.  Using the optimal
time step for time stepping yields a numerical solution with a boosted order 
of accuracy for a given finite difference scheme.

An important class of finite difference schemes for time dependent PDEs are 
those based on the method of lines.   When the method of lines approach is 
used for (\ref{eq:model_PDE}), $P$ takes a particularly simple 
form:
\beq
  P(\dx, \dt) = (\alpha \dx^r - \beta \dt^s) \dt,
  \label{eq:leading_order_error_model_PDE_MOL}
\eeq
where $r$ and $(s+1)$ are the orders of the leading spatial and temporal 
discretization errors, and $\alpha$ and $\beta$ are positive constants that 
depend on the details of the PDE and finite difference scheme.  Setting 
$P = 0$, we find that the optimal time step is given by\footnote{The $\dt = 0$ 
root of $P$ is discarded because it is not numerically meaningful.}
\beq
  \dto = \gamma \dx^{r/s},
  \label{eq:optimal_time_step}
\eeq
where $\gamma = (\alpha/\beta)^{1/s}$.  Notice that the optimal time step 
has a similar functional dependence on the grid spacing as a time step that 
satisfies a stability constraint.

\subsubsection*{\label{sec:error_analysis} 
            Error Analysis}
The form of the local error in 
(\ref{eq:leading_order_error_model_PDE_MOL}) follows from the fact 
that the method of lines first approximates the PDE as a system of ODEs in 
time by only discretizing the spatial derivatives: 
\beq
{\mathbf u}_t = F({\mathbf u}),
\label{eq:method_of_lines}
\eeq
where ${\mathbf u}$ is the vector of values of $u$ at the grid points and
$F({\mathbf u})$ is the finite difference approximation of the spatial 
derivative operator acting on ${\mathbf u}$.  The second term in the local 
error (\ref{eq:leading_order_error_model_PDE_MOL}) arises directly from 
temporal discretization errors associated with the numerical scheme used to 
integrate the system of ODEs in time.  The first term arises because $F$ is 
only an approximation of the spatial derivative operators in the PDE.  As a 
result, the system of ODEs~(\ref{eq:method_of_lines}) has an error due to 
the spatial discretization at \emph{every} point in time.  For a spatial 
discretization error of order $r$, the error in (\ref{eq:method_of_lines}) is 
$O(\dx^r)$.  Therefore, the error accumulated during each time step due to 
spatial discretization error is $O(\dt \dx^r)$ regardless of how the system of 
ODEs (\ref{eq:method_of_lines}) is solved numerically.  Because the method of 
lines is so frequently used to construct finite difference approximations for 
time dependent PDEs, we will focus solely on finite difference schemes of this 
type for the remainder of the paper to keep the discussion simple.  It is 
straightforward to extend the results we obtain to more general finite 
difference schemes.

To analyze the accuracy of the finite difference scheme 
for~(\ref{eq:model_PDE}) when the optimal time step is used, we examine the 
higher-order terms in the discretization error following the conventional 
approach (\ie ``big-O'' notation).  Elimination of the leading terms in the 
discretization error leaves a local error of the form 
$O(\dt \dx^p) + O(\dt^{q+1})$, where $p>r$ and $(q+1) > (s+1)$ are the orders 
of the spatial and temporal discretization errors, respectively, after the 
leading-order terms in the error have been eliminated.  As for the 
leading-order error (\ref{eq:leading_order_error_model_PDE_MOL}), this form of 
the local error is a direct consequence of the method of lines approach to 
discretizing the PDE.

From the local error, we can compute the global error of the numerical 
solution over extended intervals of time by using the heuristic argument
that the global error is equal to the local error divided by $\dt$.  Using 
this heuristic argument, which is rigorously justified in many texts on 
numerical methods such as~\cite{gko_book}, we find that the global error of 
the finite difference scheme using an optimal time step is 
\beq
O(\dx^p) + O(\dt^q).
\label{eq:global_error_ots}
\eeq
Observe that the global error for the finite difference scheme when an 
arbitrary time step is used has exactly the same form 
as~(\ref{eq:global_error_ots}) with $p$ and $q$ replaced by $r$ and $s$, 
respectively.  Thus, because $p > r$ and $q > s$, we see that using an optimal 
time step boosts the overall accuracy of the finite difference scheme.

It is important to emphasize that $\dx$ and $\dt$ are \emph{not} independent 
in~(\ref{eq:global_error_ots}) because the time step must be set to $\dto$ 
in order to derive the error estimate.  In other words, there is really only 
one parameter, $\dx$, that controls the numerical accuracy of the finite 
difference scheme.  This important point will be discuss in further detail in 
the following section.

\subsubsection*{Formal vs. Practical Accuracy}
For time dependent PDEs, evaluating the order of accuracy for a numerical
method is subtle because there are formally two separate orders of accuracy 
to consider\footnote{The situation is more straightforward for 
time-independent PDEs because there is usually only one order of accuracy to 
be concerned with.}:  temporal and spatial.  While it is theoretically
interesting and important to understand how the error depends on both the 
grid spacing $\dx$ and the time step $\dt$, in practice, the accuracy 
of a numerical scheme is always controlled by only one of the two numerical
parameters.

Even though spatial and temporal orders of accuracy for a numerical method
are formally separate, one will always dominate for a given choice of 
$\dx$ and $\dt$.  For example, when solving the 1D diffusion equation 
using the backward Euler method for time integration with a second-order 
central difference stencil for the Laplacian, formal analysis shows that 
the global error is $O(\dx^2) + O(\dt)$.  Since there are no 
stability constraints on the numerical parameters, the time
step and grid spacing are free to vary independently.  In this situation, the 
practical error for the method depends on the relative sizes of the time step 
and grid spacing.  When $\dt \gg \dx^2$, the practical error is 
$O(\dt)$ which means that the error in the numerical solution is 
primarily controlled by the time step.  Similarly, when 
$\dt \ll \dx^2$, the practical error is $O(\dx^2)$ so that 
the error is controlled by the grid spacing.  Finally, when 
$\dt  = O(\dx^2)$, the practical error is 
$O(\dt) = O(\dx^2)$.  In all cases, the practical error is 
primarily controlled by one of the two numerical parameters, and varying
the subdominant parameter while holding the dominant parameter fixed does 
not significantly affect the error.

When there are constraints on the numerical parameters, there is less freedom 
in choosing the controlling parameter.  For instance, if we solve the 
1D diffusion equation using a forward Euler time integration scheme with a 
second-order central difference stencil for the Laplacian, stability
considerations require that we choose $\dt = O(\dx^2)$.  
Combining this stability constraint with the formal error for the scheme
shows that the practical error is 
$O(\dx^2) + O(\dt) = O(\dx^2)$.  Thus, the accuracy of the 
method is completely controlled by the grid spacing; the temporal error can 
never dominate the spatial error.

Numerical methods for time dependent PDEs are commonly designed so that 
the global error is primarily controlled by the grid spacing.  The numerical
errors introduced by the time integration scheme are typically \emph{forced}
to be subdominant to the spatial discretization errors manually or as a result
of stability constraints.  In other words, the operational procedure used 
when developing numerical methods is usually to first choose a spatial 
discretization scheme and grid spacing so that a desired level of accuracy 
is achieved and then select the time integration scheme and time step in 
such a manner that the temporal errors do not overcome the spatial errors.  
This methodology explains the tendency to evaluate numerical schemes for
time dependent PDEs predominantly on their order of accuracy with respect 
to the grid spacing.

Optimal time step selection fits into this general framework for controlling
the error of numerical methods for time dependent PDEs.  Because the optimal 
time step is a function of the grid spacing, the technique of optimal time 
step selection places an even more restrictive constraint on $\dt$ than 
stability considerations.  Thus, like numerical methods with stability 
constraints, the accuracy of a finite difference scheme using an optimal time 
step is completely controlled by the grid spacing.  For the method of lines, 
we know that $\dto = O(\dx^{r/s})$, so the global error derived in the 
previous section reduces to 
\beq
O \left( \dx^{\min(p,rq/s)} \right).
\label{eq:global_error_ots_simplified}
\eeq

\subsubsection*{Stability Considerations}
Thus far, we have not considered the issue of numerical stability of finite
difference schemes when the time step is set to $\dto$.  For explict time 
integration schemes, OTS selection loses its value when the optimal time step 
is greater than the smallest stable time step.  In this situation, we can 
usually switch to an implicit (or semi-implicit) time integration scheme with 
the same order of accuracy for which the optimal time step is stable.  Note 
that a new optimal time step may need to be calculated for the modified finite
difference sceme.

% TABLE SHOWING COMPUTATIONAL COSTS DISCUSSED IN SECTION
% sec:computational_performance.  PLACED HERE TO SPREAD TABLES OUT
\begin{table}[!b]
\caption{\label{tab:comp_perf_vs_err} 
   Computational cost of various finite difference schemes for the 1D 
   diffusion equation as a function of the global numerical error $e$.
   For all of the schemes, the standard second-order central difference 
   stencil is used to discretize the Laplacian, and the numerical 
   solution is assumed to be computed over a fixed time interval.  
}
\begin{minipage}{\textwidth}
\begin{center} \footnotesize
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|l|c|c|c|c|}
  \hline
  {\bf Numerical Scheme} & $\dx$ 
  & $\dt$\footnote{The 
   time step for the forward/backward Euler and Crank-Nicholson schemes 
   are $O(\dx^2)$ and $O(\dx)$, respectively.  Note that the 
   time step for the backward Euler scheme is chosen so that the temporal 
   error is subdominant to the spatial error.}
  & {\bf Memory}\footnote{The memory requirement for each scheme is estimated 
    to be $O(1/\dx)$.} 
  & {\bf Compute Time}\footnote{The computation time is approximated as 
    $O\left( (1/\dx) (1/\dt) \right)$, which assumes that we are using an 
    efficient $O(N) = O(1/\dx)$ algorithm to invert the matrices that 
    arise for the implicit methods.}  \\
  \hline 
  Forward Euler    & $O\left( e^{1/2} \right)$ 
                   & $O\left( e \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-3/2} \right)$ \\
  Backward Euler   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-3/2} \right)$ \\
  Crank-Nicholson  & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-1} \right)$ \\
  Forward Euler with OTS  & $O\left( e^{1/4} \right)$ 
                   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{-1/4} \right)$ 
                   & $O\left( e^{-3/4} \right)$ \\ 
  \hline
\end{tabular}
\end{center}
\end{minipage}
\end{table}

\subsubsection*{\label{sec:computational_performance} 
                Computational Performance}
For many problems, the optimal time step is on the order of the time step 
allowed by the stability constraint for an explicit method.  
As a result, using the optimal time step to advance the finite
difference scheme in time may appear to be overly restrictive and lead to poor 
computational performance.  However, these apparent drawbacks are more than 
compensated by the boost in the order of accuracy.  For example, when solving 
the 1D diffusion equation using a forward Euler time integration scheme with a 
second-order central difference stencil for the Laplacian, we obtain a 
fourth-order scheme when $\dt$ is set equal to the optimal time step.  As
Table~\ref{tab:comp_perf_vs_err} shows, this boost in accuracy leads to a
scheme that is computationally cheaper than traditional methods that use 
the same low-order finite difference stencils.  The performance gain for 
problems in higher spatial dimensions is even more impressive.  As we can see 
in Table~\ref{tab:comp_perf_vs_dim}, the gap between the performance of a
simple forward Euler scheme with the optimal time step and the Crank-Nicholson
method grows with the spatial dimension of the problem.

\begin{table}[!t]
\caption{\label{tab:comp_perf_vs_dim}
   Computational cost of the forward Euler scheme with optimal time step and 
   the Crank-Nicholson method for solving the diffusion equation as a function 
   of the global numerical error $e$.
   For both schemes, second-order, isotropic stencils are used to discretize 
   the Laplacian.  For a problem in $d$ space dimensions, the memory 
   requirement and computation time for each scheme are estimated to be 
   $O \left(1/\dx^d \right)$ and 
   $O\left( \left(1/\dx^d \right) (1/\dt) \right)$, respectively, which 
   assumes that we are using an efficient 
   $O(N) = O \left( 1/\dx^d \right)$ algorithm to invert the matrices that 
   arise for the Crank-Nicholson method.
}
\begin{minipage}{\textwidth}
\begin{center} \footnotesize
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|c|c|c|}
  \hline
  & \multicolumn{2}{|c|}{{\bf Forward Euler with OTS}\footnote{For the forward 
    Euler scheme, the optimal time step is $\dto = \dx^2/6$.}}
  & \multicolumn{2}{|c|}{\bf Crank-Nicholson} \\
  \cline{2-3} \cline{4-5} 
    {\bf Dimensions} & {\bf Memory} & {\bf Compute Time} 
  & {\bf Memory} & {\bf Compute Time} \\
  \hline 
  $1$ & $O\left( e^{-1/4} \right)$ 
      & $O\left( e^{-3/4} \right)$ 
      & $O\left( e^{-1/2} \right)$ 
      & $O\left( e^{-1} \right)$ \\ 
  $2$ & $O\left( e^{-1/2} \right)$ 
      & $O\left( e^{-1} \right)$ 
      & $O\left( e^{-1} \right)$ 
      & $O\left( e^{-3/2} \right)$ \\ 
  $3$ & $O\left( e^{-3/4} \right)$ 
      & $O\left( e^{-5/4} \right)$ 
      & $O\left( e^{-3/2} \right)$ 
      & $O\left( e^{-2} \right)$ \\
  $d$ & $O\left( e^{-d/4} \right)$ 
      & $O\left( e^{-(d+2)/4} \right)$ 
      & $O\left( e^{-d/2} \right)$ 
      & $O\left( e^{-(d+1)/2} \right)$ \\ 
  \hline 
\end{tabular}
\end{center}
\end{minipage}
\end{table}

\subsection{\label{sec:ots_general_1d_pdes} 
            OTS Selection for General PDEs in One Space Dimension} 
In the previous section, we analyzed the technique of optimal time step 
selection in the context of a simple model PDE in one space dimension.  
For more general PDEs, the approach of carefully choosing the time step to 
eliminate leading order terms in the error continues to be useful, but it 
needs to be augmented to achieve the same improvement in the order of accuracy.
In addition to using an optimal time step, correction terms may need to be 
added to the finite difference scheme.  In this section, we discuss these 
correction terms in the context of general constant coefficient, linear PDEs 
and semilinear PDEs.

\subsubsection*{\label{sec:ots_linear_pde} 
            General Constant Coefficient, Linear PDEs} 
Let us consider a general constant coefficient, linear PDE of the form
\beq
  \frac{\partial u}{\partial t} = 
  \sum_{k=1}^n A_k \frac{\partial^k u}{\partial x^k} + f(x,t)
  \label{eq:linear_PDE}.
\eeq
Even if we use the PDE to relate terms in the discretization error, it is
unlikely that all of the dominant terms in the error can be combined into 
a single term.   As a result, the leading-order discretization error is
generally a sum of several terms:
\bea
  \dt\left[ (L u) (\alpha \dx^r - \beta \dt^s) 
  + \sum_k (L_k u) \dx^r 
  + \sum_k (L'_k u) \dt^s 
  + (G f) \dt^s \right]
  \label{eq:leading_order_error_linear_PDE_MOL}
\eea
where $L$, $u$, $\alpha$, $\beta,$ $r$, and $s$ are defined as in
(\ref{eq:leading_order_error_model_PDE_general}) and
(\ref{eq:leading_order_error_model_PDE_MOL}), 
$L_k$ and $L'_k$ are the linear spatial differential operators associated 
with the leading-order terms in the discretization error that are not involved 
in the choice of the optimal time step, and $G$ is a spatio-temporal 
differential operator that acts on source term $f$.  
Note that neither $L_k$ nor $L'_k$ involve any temporal differential 
operators because any that arise can be eliminated using the 
PDE~(\ref{eq:linear_PDE}).  Also, there is no explicit spatial discretization 
error of order $r$ that originates from the source term because it is not 
differentiated with respect to space when the semi-discrete 
equations~(\ref{eq:method_of_lines}) are derived from the PDE.

From~(\ref{eq:leading_order_error_linear_PDE_MOL}), it is clear that setting 
$\dt = \dto$ is not enough to completely eliminate the 
leading-order error.  To eliminate the vestigial errors, we need to add a few 
correction terms to the numerical scheme in exactly the same spirit used in 
the design of the modified nine-point formula for the Poisson 
equation~\cite{iserles_book}.  The required correction terms are precisely 
those that cancel out the second, third, and fourth terms in 
(\ref{eq:leading_order_error_linear_PDE_MOL}).  Fortunately, it is 
straightforward to compute and incorporate these terms into the numerical 
scheme.  At each time step, the second and third terms in 
(\ref{eq:leading_order_error_linear_PDE_MOL}) can be approximated using
finite differences applied to the solution $u$ at the current time.
The last term can be computed using either a finite difference approximation 
or calculating it analytically.  With the addition of these correction terms, 
the leading-order discretizaton error is completely eliminated, and the
global error can be determined using the analysis in 
Section~\ref{sec:error_analysis}.

It is important to point out that the finite difference stencils used to 
approximate the vestigial errors must be chosen carefully.  Special attention 
is required to ensure that the finite difference stencils are of sufficiently 
high-order.  Otherwise, we will lose the accuracy gained from using an optimal 
time step and adding the correction terms to the finite difference scheme.  
Since the specific form of the correction terms are PDE-dependent, we defer 
further discussion to Sections~\ref{sec:applications_1d} 
and~\ref{sec:applications_multidim} where we apply optimal time step selection 
to several specific PDEs.

\subsubsection*{Semilinear PDEs}
Semilinear PDEs are an important class of PDEs that have the form
\beq
  \frac{\partial u}{\partial t} = A(x,t) \frac{\partial^n u}{\partial x^n}
  + F \left( \frac{\partial^{n-1} u}{\partial x^{n-1}},
      \frac{\partial^{n-2} u}{\partial x^{n-2}}, \ldots,
      \frac{\partial u}{\partial x}, u \right)
  + f(x,t)
  \label{eq:semilinear_PDE}
\eeq
where $F$ is an arbitrary function of the lower-order spatial derivatives
of $u$ and $f(x,t)$ is a source term.  While the coefficient on the 
leading-order spatial derivative is generally allowed to be function of 
space and time, optimal time step selection can only be applied to PDEs where
$A(x,t)$ is a constant.  Because the PDE is linear in the leading-order 
spatial derivative, it is natural to construct a finite difference scheme that 
treats $F$ in an explicit manner and lets stability considerations determine 
whether the $A \frac{\partial^n u}{\partial x^n}$ term is handled explicitly 
or implicitly.  Assuming that sufficiently high-order stencils are used to 
compute $F$, the leading-order discretization error for such a finite 
difference scheme is given by
\bea
  \dt \left[ (L u) (\alpha \dx^r - \beta \dt^s) 
  + \left( H u \right) \dx^r 
  + \left( H' u \right) \dt^s
  + (G f) \dt^s \right]
  \label{eq:leading_order_error_semilinear_PDE_MOL}
\eea
where $L$, $u$, $\alpha$, $\beta,$ $r$, and $s$ are defined as in
(\ref{eq:leading_order_error_model_PDE_general}) and
(\ref{eq:leading_order_error_model_PDE_MOL}), $H$ and $H'$ are spatial 
differential operators (potentially nonlinear) associated with the 
leading-order terms in the discretization error that arise from the nonlinear 
term in~(\ref{eq:semilinear_PDE}), and $G$ is a spatio-temporal differential 
operator that acts on $f$.  With the discretization error written in the 
form (\ref{eq:leading_order_error_semilinear_PDE_MOL}), the optimal time 
step and correction terms can be computed by proceeding in the same manner 
as for general constant coefficient, linear PDEs.

\subsection{Boundary Conditions}
In our analysis and discussion thus far, we have completely ignored the issue 
of boundary conditions.  The choice of discretization for boundary conditions
can have a significant impact on the accuracy of the numerical solution.  In 
general, high-order discretizations of boundary conditions are required to 
ensure that errors introduced at the domain boundary do not pollute the 
numerical solution in the domain interior.  For Dirichlet boundaries, it
is sufficient to impose boundary conditions by using boundary values that have 
the same order of accuracy as the solution.  More effort is required to ensure 
that Neumann boundary conditions do not reduce the order of accuracy of the 
solution.  In general, Neumann boundary conditions require that the accuracy 
of ghost cells slightly exceed the desired order of accuracy for the 
solution.  A detailed discussion of methods for imposing boundary conditions 
is beyond the scope of this paper.  We refer interested readers to the 
thorough analysis presented in~\cite{gko_book}.  To avoid complicating the 
discussion in Sections~\ref{sec:applications_1d} 
and~\ref{sec:applications_multidim}, Neumann boundary conditions are imposed 
by taking advantage of knowledge of the analytical solution.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Manuscript started: 2008/09/0?
% First completed draft: 2008/??/??, Vitamin D, Inc.
% Submission draft: 2008/??/??, Vitamin D, Inc.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\title{High-Order Accurate Finite Difference Schemes \\
       for Systems of PDEs via Optimal Time Step Selection}

@@

\begin{abstract}
place holder...
\end{abstract}

\begin{keywords}
optimal time step, finite difference schemes, high-order accurate numerical 
methods, time dependent PDEs, systems of PDEs
\end{keywords}

\begin{AMS}
65-02, 65M06, 65M12, 65M20, ??
\end{AMS}

@@

\section{\label{sec:summary} Summary}

\section*{Acknowledgments}
The author gratefully acknowledges the support of Vitamin D, Inc.
and the Institute for High-Performance Computing (IHPC) in Singapore. 
The author would like to thank K.-H. Chiam, ?? 
for helpful suggestions on the manuscript.  
The author particularly thanks K.-H. Chiam for pointing him to the 
mathematical biology literature for interesting examples of reaction-diffusion 
systems.

\bibliography{OTSPDE-Systems}
